Fast Sorting on Modern Computers

### 2005-05-31 paper/docs/merge.tex
have to rewrite a section of the merges code. starting now

The next problem considered is that of temporal locality. Once a key is loaded
into the array, it's only used once per merge step. If the data does not fit in
half the cache, then conflict misses will occur, and no temporal reuse will
occur at all. The solution is to apply a technique used in compilers called
\n{tiling}. Using this technique, the number of keys which fit in the cache will
be fully sorted before moving on to the next set. When all the sets have been
sorted in the manner, they are all merged together as usual.

@@

The final improvement was to fully sort keys in the level 1 cache, before
continuing to level 2 cache as in the tiled mergesort. The expected improvement
is not as significant as in the case of the level 2 cache, since is the penalty
for level 1 misses is not great. For this reason, this sort is called Double
Tiled Mergesort instead of Tiled Mergesort, however, it is referred to as both
in this report.

@@

Double Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists sized $C/2$. The second phase attempts to improve
temporal locality by merging these lists together in a single merge step. There
are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
search across each of these keys, maintaining a list of indices to the smallest
unused key from each list. When there is a small number of lists, this isn't a
problem, however, when the lists grow big, the overhead involved in this is
substantial. It is necessary to check each of the lists, and each key searched
would cause a cache miss, since each array segment is aligned, and map to
the same cache block.

To avoid this, a priority queue is used. The smallest key from each list is
added to the queue, and then put into the final target array. This reduces the
problem of the search overhead, but does not reduce the problem of
\n{thrashing}\footnote{Thrashing occurs when two arrays are mapped to the same
cache block, and the constant access of one followed by the other causes every
access to be a miss.}.  To avoid this, an entire cache line is added at a time
to the priority queue. There will still be conflict misses, but there will be no
thrashing since key overwritten in the cache will not be needed again.

@@

Using a sorted list instead of LaMarca's tagging leads to a variety of problems.
The reason the sorted list is used is that no details of the tagging is
provided. The sorted list is $k$ keys long, and only needs to be properly sorted
once. The next time it needs to be sorted is when a new line is added, but in
this case only one key is out of place, and can be put in place with a single
insertion sort run. This is linear at worst, and has excellent branch prediction
properties, as shown earlier.

When a list is fully consumed, the next keys inserted will be from the next
list. This could be fixed with a bounds check, however, since the lists are
bitonic, the next keys inserted are going to be the largest keys from the next
array. These will be added to the heap, and take their place at the bottom,
without harmful effect. A harmful effect is possible, though, when lots of keys 
with the same value are in the array. All these keys would be inserted at a
similar time, and when the first one is taken off the heap, a list would add
eight keys, even though it had just added eight. In this manner, the size of the
heap could be overrun. With random keys in the array, this is unlikely, and so
this is not addressed.

@@

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8, so 2048 keys can be sorted in the level 1 cache.
The number of keys to be sorted in the level 2 cache is 262144.  When arrays
were pre-sorted, they were pre-sorted to a size four, or eight if the number of passes
was going to be odd. These were reversed for the double multi-mergesort; since
it does a k-way merge into the source array, the sorted cache sized segments
need to be in the auxiliary array. An 8-heap was used for the multi-mergesorts.

@@

The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
The reason for this is not clear.

>>>>

The next problem considered is that of temporal locality. Once a key is loaded
into the array, it's only used once per merge step. If the data does not fit in
half the cache, then conflict misses will occur, and no temporal reuse will
occur at all. The solution is to apply a technique used in compilers called
\n{tiling}. Using this technique, an entire cache sized portion of the array will
be fully sorted before moving on to the next set. When all the sets have been
sorted in the manner, they are all merged together as usual.

@@

The final improvement was to fully sort keys in the level 1 cache, before
continuing to level 2 cache as in the tiled mergesort. The expected improvement
is not as significant as in the case of the level 2 cache, since the penalty
for level 1 misses is not great. For this reason, this sort is called Double
Tiled Mergesort instead of Tiled Mergesort, however, it is referred to as both
in this report.

@@

Double Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists sized $C/2$. The second phase attempts to improve
temporal locality by merging these lists together in a single merge step. There
are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
search across each of these keys, maintaining a list of indices to the smallest
unused key from each list. When there is a small number of lists, this isn't a
problem, however, when the number of lists increases significanty, the overhead
involved in this is substantial. It is necessary to check each of the lists, and
each key searched would cause a cache miss, since each array segment is aligned,
and map to the same cache block.

To avoid this, a priority queue is used. The smallest key from each list is
added to the queue, and then put into the final target array. This reduces the
problem of the search overhead, but does not reduce the problem of
\n{thrashing}\footnote{Thrashing occurs when two arrays are mapped to the same
cache block, and the constant access of one followed by the other causes every
access to be a miss.}.  To avoid this, an entire cache line is added at a time
to the priority queue. There will still be conflict misses, but there will be no
thrashing since the keys overwritten in the cache will not be needed again.

@@

Using a sorted list instead of LaMarca's tagging leads to a variety of problems.
The reason the sorted list is used is that no details of the tagging is
provided. The sorted list is $k$ keys long, and only needs to be properly sorted
once. The next time it needs to be sorted is when a new line is added, but in
this case only one key is out of place, and can be put in place with a single
insertion sort run. This is linear at worst, and will have only 1 branch
misprediction.

When a list is fully consumed, the next keys inserted will be from the next
list. This could be fixed with a bounds check, however, since the lists are
bitonic, the next keys inserted are going to be the largest keys from the next
array. These will be added to the heap, and take their place at the bottom,
without harmful effect. A harmful effect is possible, though, when lots of keys 
with the same value are in the array. All these keys would be inserted at a
similar time, and when the first one is taken off the heap, a list would add
eight keys, even though it had just added eight. In this manner, the size of the
heap could be overrun. With random keys over a large range, as in our tests,
this is unlikely, and so this is not addressed.

@@

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8k, so 2048 keys can be sorted in the level 1 cache.
262144 keys can be sorted in the level 2 cache. When arrays were pre-sorted,
they were pre-sorted to a size four, or eight if the number of passes was going
to be odd. These were reversed for the double multi-mergesort; since it does a
k-way merge into the source array, the sorted cache sized segments need to be in
the auxiliary array. An 8-heap was used for the multi-mergesorts.

@@

%TODO look over lamarcas results
The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
The reason for this is not clear.
